Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-05-06 paper/main.tex
Updates from Overleaf

Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.

@@

This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.\\

@@

Predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 

@@

Here we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimization of protein level expression. RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
We have used Gaussian Process-Upper Confidence Bound and multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS \cite{desautels2012parallelizing, Rasmussen2004}.

@@

In the subsequent rounds, designs where recommended by the algorithm based on the data obtained in previous rounds. 

@@

The results were then fed back to the algorithm for it to recommend the next round of designs.\\

@@

Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised. Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase. 
In LEARN phase we use the Gaussian Process regression algorithm to predict the behaviour of different RBS sequences and in the DESIGN phase we use the multi-armed Bandit algorithm to recommend sequences to be Built and Tested next. 
Conventionally, the Design phase is considered the beginning of the cycle, however we will start description of our results from the Learn phase, which we find a more natural starting point for a machine learning centred workflow.\\

@@

\subsection{DESIGN of the genetic device}
There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992,SHULTZABERGER2001}.
In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\

>>>>

Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.

@@

This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.\\

@@

Reported predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 

@@

Here we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimisation of associated protein level expression. 
RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
In this work We have used Gaussian Process Regression and Upper Confidence Bound multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS \cite{desautels2012parallelizing, Rasmussen2004}.

@@

In the subsequent rounds, designs where recommended by the algorithm based on the data obtained in the previous rounds. 

@@

The results were then fed back to the algorithm for it to recommend the next round of designs.
This way, were able to build an extensive, reliable library of novel RBSs with diverse sequences.
At the same time we were able to discover new RBS sequences with very high TIRs.\\

@@

Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised. Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase. 
In LEARN phase we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences and in the DESIGN phase we use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next. 
The following sections will discuss our machine learning workflow divided into specific parts of the DBTL cycle.\\

@@

\subsection{DESIGN of the genetic device}
There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992,SHULTZABERGER2001}.
In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
Additionally, there is a consensus RBS core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\
